Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-05 Manuscript/manuscript.Rmd
manuscript edits

By isolating bins that overlapped with flanked TAD boundaries we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. To  get a hollisitic view of how some annotations tended to occur together at boundaries, we computed the average pairwise correlations between count overlaps, count percents, and $log_{2}$ distances of each genomic annotation with bins that overlapped with flanked TAD boundaries. Clustered correlograms at the 10 kb resolution, for both cell lines, showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Figure 7). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. Similar correlations were exhibited across the remainder of resolutions, for both cell lines (Supplementary Figures 2-4). The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed predictor-type and resolution specific data after implementing each of the resampling techniques within each cell line.

@@

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions for each cell line (Figure 7 (A)). Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

@@

Comparisons of the effects of the various resampling techniques were then made across resolutions, for each cell line. Results showed that performances drastically improved among the composite metrics (Figure 7 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.

@@

# DISCUSSION

@@

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.

>>>>

By isolating bins that overlapped with flanked TAD boundaries we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. To  get a hollisitic view of how some annotations tended to occur together at boundaries, we computed the average pairwise correlations between count overlaps, count percents, and $log_{2}$ distances of each genomic annotation with bins that overlapped with flanked TAD boundaries. Clustered correlograms at the 10 kb resolution, for both cell lines, showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Figure 7). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. Similar correlations were exhibited across the remainder of resolutions, for both cell lines (Supplementary Figures X-Y). The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed predictor-type and resolution specific data after implementing each of the resampling techniques within each cell line.

@@

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions for each cell line (Figure 8 (A)). Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

@@

Comparisons of the effects of the various resampling techniques were then made across resolutions, for each cell line. Results showed that performances drastically improved among the composite metrics (Figure 8 (B-D)). Models implemented on randomly under-sampled and SMOTE-sampled data performed similarly, while performances on randomly over-sampled data lagged behind, yet still outperformed models with no data re-sampling. Overall, there appeared the be a downward trend in model performance as resolutions decreased, across both cell lines.

@@

<!-- IN PROGRESS
# DISCUSSION

@@

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
-->
